---
title: "stuff"
author: "Matt Jones"
date: "7/19/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Many people write R code as a single, continuous stream of commands, often drawn
from the R Console itself and simply pasted into a script.  While any script 
brings benefits over non-scripted solutions, there are advantages to breaking
code into small, reusable modules.  This is the role of a `function` in R.  In
this lesson, we will review the advantages of coding with functions, practice 
by creating some functions and show how to call them, and then do some exercises
to build other simple functions.

## Leaning outcomes

- Learn why we should write code in small functions
- Write code for one or more functions
- Document functions to improve understanding and code communication

## Why functions?

In a word:

- DRY: Don't Repeat Yourself

By creating small functions that only one logical task and do it well, we quickly 
gain:

- Improved understanding
- Reuse via decomposing tasks into bite-sized chunks
- Improved error testing


## Temperature conversion

Imagine you have a bunch of data measured in Fahrenheit and you want to convert
that for analytical purposes to Celsius.  You might have an R script
that does this do you.

```{r}
airtemps <- c(212, 30.3, 78, 32)
celsius1 <- (airtemps[1]-32)*5/9
celsius2 <- (airtemps[2]-32)*5/9
celsius3 <- (airtemps[3]-32)*5/9
```

Note the duplicated code, where the same formula is repeated three times.  This
code would be both more compact and more reliable if we didn't repeat ourselves.

## Creating a function

Functions in R are a mechanism to process some input and return a value.  Similarly
to other variables, functions can be assigned to a variable so that they can be used
throughout code by reference.  To create a function in R, you use the `function` function (so meta!) and assign its result to a variable.  Let's create a function that calculates
celsius temperature outputs from fahrenheit temperature inputs.

```{r}
fahr_to_celsius <- function(fahr) {
  celsius <- (fahr-32)*5/9
  return(celsius)
}
```

By running this code, we have created a function and stored it in R's global environment.  The `fahr` argument to the `function` function indicates that the function we are creating takes a single parameter (the temperature in fahrenheit), and the `return` statement indicates that the function should return the value in the `celsius` variable that was calculated inside the function.  Let's use it, and check if we got the same value as before:

```{r}
celsius4 <- fahr_to_celsius(airtemps[1])
celsius4
celsius1 == celsius4
```

Excellent.  So now we have a conversion function we can use.  Note that, because 
most operations in R can take multiple types as inputs, we can also pass the original vector of `airtemps`, and calculate all of the results at once:

```{r}
celsius <- fahr_to_celsius(airtemps)
celsius
```

This takes a vector of temperatures in fahrenheit, and returns a vector of temperatures in celsius.

## Exercise

Now, create a function named `celsius_to_fahr` that does the reverse, it takes temperature data in celsius as input, and returns the data converted to fahrenheit.  Then use that formula to convert the `celsius` vector back into a vector of fahrenheit values, and compare it to the original `airtemps` vector to ensure that your answers are correct.

```{r}
# Your code goes here
```

Did you encounter any issues with rounding or precision?

