---
title: "SQL Database Access in R"
time_slot: 60 minutes
---

## Introduction

Continuing on from our earlier lesson on SQL, we're going to learn how to work with SQL and database in R.
There are two ways we're going to go over:

1. Directly executing SQL statements against a database
2. Using `dplyr`'s seamless ability to work with SQL databases

## Learning outcomes

Students should:

- Be able to connect to an SQL database in R
- Be able to load external data in CSV format into a table in the database
- Be able to run a query against a database table in R
- Be able to run an SQL join against two database tables in R
- Become familiar with connecting `dplyr` to a database

## Lesson

### Work with SQL directly using the `DBI` package

This content is largely adapted from the [DBI homepage](http://rstats-db.github.io/DBI/).

```{r}
# Create an ephemeral in-memory RSQLite database
library(DBI)
# Create an ephemeral in-memory RSQLite database
con <- dbConnect(RSQLite::SQLite(), dbname = ":memory:")

dbListTables(con)
dbWriteTable(con, "mtcars", mtcars)
dbListTables(con)

dbListFields(con, "mtcars")
dbReadTable(con, "mtcars")

# You can fetch all results:
res <- dbSendQuery(con, "SELECT * FROM mtcars WHERE cyl = 4")
dbFetch(res)
dbClearResult(res)

# Or a chunk at a time
res <- dbSendQuery(con, "SELECT * FROM mtcars WHERE cyl = 4")
while(!dbHasCompleted(res)){
  chunk <- dbFetch(res, n = 5)
  print(nrow(chunk))
}
dbClearResult(res)

dbDisconnect(con)
```

**Exercise**: DBI and starwars

Instructions: Load the starwars dataset from the `dplyr` package into an in-memory SQLite database and execute a query against the table.

e.g. list just the Droids

```{r}
# Your code here
```

### Automatic SQL with dplyr

```{r}
library(dplyr)
library(dbplyr)

mtcars_db <- tbl(con, "mtcars")
mtcars_db

mtcars_db %>% 
  select(mpg)

mpg_query <- mtcars_db %>% 
  group_by(cyl) %>% 
  summarize(mean(mpg))
mpg_query %>% show_query()

# To get the data out of our query, we use collect()
class(mpg_query)
mpg_query %>% collect() %>% class()
```

There, that's it.
`dbplyr` automatically converts our `dplyr` code into SQL queries behind the scenes.

***Exercise: Load the `diamonds` dataset from the `ggplot2` package and calculate the average price for each carat and make a plot of the relationship:

```{r}
# Your code here
```

## Summary

- Using SQL via R only requires a couple of lines of code
- We can use `dplyr` against an SQL database for intuitive querying that is fast