---
title: R Review & Assessment
time_slot: 60 minutes"
---

## Introduction

Let's take about an hour to touch up on basic R programming skills in a survey style and then we'll move onto some challenges which should test your understanding.

## Learning Outcomes

- Refresh general R programming skills
- Test R skills against some example problems

## Assessment

Everyone will arrive to this lesson with different experiences with R.
Skill with R isn't necessarily a continuum and can instead be thought of as a set of tools.
Each particpant may start the workshop with different tools.

In order to get a better sense of what topics in R we should focus more heavily on, let's do a quick assessment.
The results will help us shape instruction so that we can ensure we're meeting everyone's needs.

**Instructions:** 

TODO: Decide if I really like/prefer a Google Form to get more quantitative

Question ideas:

### Question 1

Which of the following expressions assigns the number 2 to the variable x?

A. `x == 2`
B. `x <- 2`
C. `x - 2`
D. `x = 2`

### Question 2

What does the following expression return?

```{r}
paste("apple", "pie")
```

A. "applepie"
B. "apple, pie"
C. "apple pie"
D. An error
  
### Question 3

If x and y are both data.frames, 

```{r}
x <- data.frame(z = 1:5)
y <- data.frame(z = 6:10)
```

which of the following expressions would be a correct way to combine them into one data.frame that looks like this:

z
-
1
2
3
4
5
6
7
8
9
10

A. `join(x, y)`
B. `c(x, y)`
C. `rbind(x, y)`
D. `x + y`

### Question 4
  
Given the following data.frame,

x <- data.frame(y = 1:10)

Which expression(s) return a `data.frame` with rows where y is greater than 5 (i.e. 6 - 10)

A. `x[x$y > 5,]`
B. ``x$y > 5`
C. `x[which(x$y > 6),]`
D. `x[y > 5,]`
E. `subset(x, y > 5)`
    
### Question 5

What does the following bit of code print out?

```{r}
for (i in 1:10) {
  if (i %% 2) {
    print(i)
  }
}
```

A. An error
B. 1, 2, 3, 4, 5, 6, 7, 8, 9, 10
C. 2, 4, 6, 8, 10
D. 1, 3, 4, 7, 9

### Question 6

Given the following `data.frame`:

`x <- data.frame(x = LETTERS)`

What is the result of this expression?

`class(x$x)`

A. factor
B. numeric
C. logical
D. character

### Question 7

Given a list of the letters in the English alphabet defined by:

`x <- list("y" = LETTERS)`

What does the following expression result in?

`length(x["y"])`

A. 0
C. An error
C. 26
D. 1

### Question 8
TODO

### Question 9
TODO

### Question 10

If package `foo` contains a function `bar` and package `baz` also contains a function `bar`, defined as such:

```
# in package foo
bar <- function() {
  return "buzz"
}

# in package baz
bar <- function() {
  return "woof"
}
```

What output, if any, does the following line of code produce?

```
library(foo)
library(baz)
bar()
```

## R overview

Based on previous R training: https://github.nceas.ucsb.edu/Training/R-intro.
Instructor will go over these live with the classroom.

- basic syntax
  - comparison
- variables / assignemtn
    - assignment
- types
    - vectors
        - character vecs / strings
    - lists
    - vectors vs. lists
        - touch on the fact that vectors are for one type and lists are for multi
        - what does this do? `r c(1, "asdf", 2)`
    - data.frames
- data import/export (e.g. read.csv)
  - read.csv
  - read.table
  - readLines
  - save/load
- data manip
    - subset()
    - data.frame [] notation
- modeling
- sample / runif / rnorm
  - sampling a data.frame
- common util functions
    - paste (combine things)
    - table
    - summary
    - str
    - seq
    - nchar
- iteration
    - for loops
        - when do we use for loops? when we're doing for each X where X can be year, site, etc.
    - whie loops
- apply-family
  - sapply
  - lapply
- functions: We'll skip this as it will be taught in another lesson

More advanced topics to cover if time allows:

- Writing faster code
  - What is 'vectorization'?
  - Don't grow variables
- Debugging
  - traceback()
  - browser()
  - breakpoints in RStudio
  
## Call your shot

An excellent way to really learn a programming language is to call out what the result of running some expression will be *before* you run it.
Afterwards, you can compare your expectation with what actually happened.

Here are code chunks with a series of expressions.
Try to predict what **the final expression* does before running the entire chunk:


```{r}
x <- 2
x ^ 2
```

```{r}
x <- 1; y <- 2; x + y;
```

```{r}
x <- "hello"
y <- "world"
paste(x, y)
```

```{r}
x <- list(1, 2, 3)
y <- list(4, 5, 6)
z <- c(x, y)
length(z)
```

```{r}
x <- data.frame(x = 1:6)
y <- data.frame(x = 1:7)
z <- rbind(x, y)
nrow(z)
```


```{r}
x <- NA

if (is.na(x)) {
  print("foo")
} else {
  print("bar")
}
```

```{r}
numbers <- seq(1, 10)

for (number in numbers) {
  if (number %% 2){
    print(number)
  }
}
```

```{r}
x <- 10

while (x >= 0) {
  print(x)
  
  if (x == 5) {
    break
  }
  
  x <- x - 1
}
```

```{r}
x <- c(1, "2", 3)
class(x)
```

```{r}
x <- list(1, 2, 3)
# todo: something with lists
```

```{r}
x <- data.frame(letter = LETTERS)
class(x[1,])
```

```{r}
x <- data.frame(letter = LETTERS)
class(x[1, 1, drop = FALSE])
```

```{r}
x <- data.frame(x = 1)
y <- data.frame(x = 2)
z <- rbind(x[1,1], y[1,1])
class(z)
```
```{r}
rep(TRUE, 5) & rep(FALSE, 5)
```

```{r}
rep(TRUE, 5) && rep(FALSE, 5)
```

```{r}
x <- c(1, 2, 3)
y <- c("A", "B", "C")
z <- c(x, y)
class(z)
```

```{r}
x <- c(1, 2, NA, 4, NA)
length(is.na(x))
```

```{r}
x <- c(1, NA, 3)
y <- c(NA, 2, NA)
all(is.na(x + y))
```

TODO: Example where factors trip you up

## Summary

By the end of this lesson, you should have feel touched up on your general R skills and you also should have seen some of the trickier parts of R.
Hopefully having seen the trickier parts of R will help later on down the road.

## Resources

Other good resources:

- One of our lessons on R: https://github.nceas.ucsb.edu/Training/R-intro
- Great, free book: http://r4ds.had.co.nz/
- DataCamp's R intro, https://www.datacamp.com/courses/free-introduction-to-r
Sources:
- Used https://www.rstudio.com/resources/cheatsheets/ to make sure I wasn't missing any basic stuff