---
title: R Review & Assessment
time_slot: 60 minutes"
---

## Introduction

Let's take about an hour to take a quick assessment, touch up on basic R programming skills in a survey style, and then we'll move onto some challenges which should test your understanding if we have time.

## Learning Outcomes

- Assess where everyone is with R so we can shape the curricula
- Refresh general R programming skills
- Test R skills against some example problems

## Assessment

Everyone will arrive to this lesson with different experiences with R.
Skill with R doesn't necessarily exist a continuum and can instead be thought of as a set of tools.
Each particpant may start the workshop with different tools.

In order to get a better sense of what topics in R we should focus more heavily on, let's do a quick assessment.
The results will help us shape instruction so that we can ensure we're meeting everyone's needs.

**Instructions:** 

Answer the following 5 questions to the best of your knowledge and write down your answers.

### Question 1

Which of the following expressions assigns the number 2 to the variable x?

Choose one or more:

- A. `x == 2`
- B. `x <- 2`
- C. `x - 2`
- D. `x = 2`

Your answer:

### Question 2

What does the following expression return?

```{r}
paste("apple", "pie")
```

Choose one:

- A. "applepie"
- B. "apple, pie"
- C. "apple pie"
- D. An error

Your answer:

### Question 3

What does the following expression return?

```{r}
max(abs(c(-5, 1, 5)))
```

Choose one:

- A. -5
- B. 1
- C. 5
- D. An error

Your answer:

### Question 4

If x and y are both data.frames defined by:

```{r}
x <- data.frame(z = 1:2)
y <- data.frame(z = 3)
```

which of the following expressions would be a correct way to combine them into one data.frame that looks like this:

```
z
-
1
2
3
```

(i.e. one column with the numbers 1, 2, and 3 in it)

Choose one or more:

- A. `join(x, y)`
- B. `c(x, y)`
- C. `rbind(x, y)`
- D. `x + y`

Your answer:

### Question 5
  
Given the following data.frame,

x <- data.frame(y = 1:10)

Which expression(s) return a `data.frame` with rows where y is greater than 5 (i.e. 6 - 10)

Choose one or more:

- A. `x[x$y > 5,]`
- B. ``x$y > 5`
- C. `x[which(x$y > 6),]`
- D. `x[y > 5,]`
- E. `subset(x, y > 5)`
    
## R overview

Based on previous R training: https://github.nceas.ucsb.edu/Training/R-intro.
Instructor will go over these live with the classroom, running 

- Basic syntax
  - Variables & assignemnt
  
### The assignment operator, `<-`

One of the things we'll do all the time is save some value to a variable.
Here, we save the word "apple" to a variable called `fruit`

```{r}
fruit <- "apple"
fruit
```

Notice the last line with just `fruit` on it.
Typing just the variable name in just prints the value to the Console.

R has a flexible syntax.
The following two lines of code are identical to the above one.

```{r}
fruit<-"apple"
fruit    <-     "apple"
```

### R as a calculator: `+ - * / > >= %% %/%` etc
  
```{r}
2+2
2 * 3
2 ^ 3
5/2
```

Comparison:

```{r}
2 == 1
2 == 2
3 > 2
2 < 3 # Same as above
"apple" == "apple"
"apple" == "pair"
"pair" == "apple" # Order doesn't matter for ==
```


### Types of variables

#### Vectors

When we run a line of code like this:

```{r}
x <- 2
```

We're assigning 2 to a variable `x`.
`x` is a variable but it is also a "numeric vector" of length 1.

```{r}
class(x)
length(x)
```

Above, we ran two function: `class` and `length` on our variable `x`.
Running functions is a very common thing you'll do in R.
Every function has a name, following by a pair of `()` with something inside.

We can make a numeric vector that is longer like so:

```{r}
x <- c(1, 2, 3) # Use the `c` function to put things together
```

Notice we can also re-define a variable at a later point just like we did above.

```{r}
class(x)
length(x)
```

R can store much more than just numbers though.
Let's start with strings of characters, which we've already seen:

```{r}
fruit <- "apple"
class(fruit)
length(fruit)
```

Depending on your background, you may be surprised that the result of running `length(fruit)` is 1 because "apple" is five characters long.

It turns out that `fruit` is a character vector of length one, just like our numeric vector from before.
To find out the number of characters in "apple", we have to use another function:

```{r}
nchar(fruit)
nchar("apple")
```

Let's make a character vector of more than length one and take a look at how it works:

```{r}
fruits <- c("apple", "banana", "strawberry")
length(fruits)
nchar(fruits)
fruits[1]
```

Smushing character vectors together can be done with `paste`:

```{r}
paste("key", "lime", "pie")
```

#### Lists

Vectors and lists look similar in R sometimes but they have very different uses:

```{r}
c(1, "apple", 3)
list(1, "apple", 3)
```

#### data.frames
Most of the time when doing analysis in R you will be working with `data.frames`.
`data.frames` are tabular, with column headings and rows of data, just like a CSV file.

We create new `data.frames` with a relevantly-named function:

```{r}
mydata <- data.frame(site = c("A", "B", "C"),
                     temp = c(20, 30, 40))
mydata
```

Or we can read in a CSV from the file system and turn it into a `data.frame` in order to work with it in R:

```{r}
mydata <- read.csv("data.csv")
mydata
```

We can find out how many rows of data `mydata` has in it:

```{r}
nrow(mydata)
```

We can return just one of the columns:

```{r}
mydata$type
unique(mydata$type)
```
sort

If we want to sort `mydata`, we use the `order` function (in kind of a weird way):

```{r}
mydata[order(mydata$type),]
```

Let's break the above command down a bit.
We can access the individual cells of a `data.frame` with a new syntax element: `[` and `[`:

```{r}
mydata[1,] # First row
mydata[,1] # First column
mydata[1,1] # First row, first column
mydata[c(1,5),] # First and second row
mydata$type # Column named 'type'
```

So what does that `order` function do?

```{r}
?order # How to get help in R!
order(c(1, 2, 3))
order(c(3, 2, 1))
order(mydata$type)
```

So `order(mydata$type)` is returning the rows of `mydata`, by row number, in sorted order.

We can also return just certain rows, based upon criteria:

```{r}
mydata[mydata$type == "fruit",]
mydata$type == "fruit"
```

In this case, instead of indexing the rows by number, we're using TRUEs and FALSEs.

**Exercise:** Subset `mydata` to the vegetables instead of the fruit

```{r}
# Your code here
```

Another handy way to subset `data.frame` is with the `subset` function:
    
```{r}
subset(mydata, type == "fruit") # Equivalent to mydata[mydata$type == "fruit",]
```

There are a lot of useful functions to help us work with `data.frame`s:

```{r}
str(mydata)
summary(mydata)
```

Our `data.frame`s won't always be so small as this example one.
Let's look at a larger one:

```{r}
library(ggplot2)
data("diamonds")
diamonds
```

**Exercise:** How many rows does diamonds have in it? How many columns?

We can look at the first few rows with `head`, just like on the command line:

```{r}
head(diamonds)
```

or the last few:

```{r}
tail(diamonds)
```
So far, this has probably been a bit boring.
Let's do something interesting and also something that R is very good at: Plotting and modeling!

Let's plot the relationship between diamond price and carat:

```{r}
plot(price ~ carat, data = diamonds)
```

The above syntax, `price ~ carat` uses a `response ~ predictor` form or `y ~ x`.

We can also fit a linear model to the same relationship:

```{r}
mod <- lm(price ~ carat, data = diamonds)
```

Above, we saved our linear model to a variable named `mod` so we can use it later.
We can look at the result of model fitting with `summary`.

```{r}
summary(mod)
```

And we can also plot the line of best fit on the scatterplot:

```{r}
plot(price ~ carat, data = diamonds)
abline(mod$coefficients[[1]], mod$coefficients[[2]], col = "red", lwd = 5)
```

### More advanced topics to cover if time allows:

- Modeling
  - `sample` / `runif` / `rnorm`
    - sampling a data.frame
- Common util functions
    - `table`
- Iteration
    - `for` loops
        - When do we use for loops?
    - `while` loops
- apply-family
  - `sapply`
  - `lapply`
- functions: We'll skip this as it will be taught in another lesson
- Writing faster code
  - What is 'vectorization'?
  - Don't grow variables
- Debugging
  - traceback()
  - browser()
  - breakpoints in RStudio
  
## Call your shot

An excellent way to really learn a programming language is to call out what the result of running some expression will be *before* you run it.
Afterwards, you can compare your expectation with what actually happened.

Here are code chunks with a series of expressions.
Try to predict what **the final expression* does before running the entire chunk and add a note if you got one wrong.

```{r}
x <- 2
x ^ 2
```

```{r}
x <- 1; y <- 2; x + y;
```

```{r}
x <- "hello"
y <- "world"
paste(x, y)
```

```{r}
x <- list(1, 2, 3)
y <- list(4, 5, 6)
z <- c(x, y)
length(z)
```

```{r}
x <- data.frame(x = 1:6)
y <- data.frame(x = 1:7)
z <- rbind(x, y)
nrow(z)
```


```{r}
x <- NA

if (is.na(x)) {
  print("foo")
} else {
  print("bar")
}
```

```{r}
numbers <- seq(1, 10)

for (number in numbers) {
  if (number %% 2){
    print(number)
  }
}
```

```{r}
x <- 10

while (x >= 0) {
  print(x)
  
  if (x == 5) {
    break
  }
  
  x <- x - 1
}
```

```{r}
x <- c(1, "2", 3)
class(x)
```

```{r}
x <- list(1, 2, 3)
lapply(x, cumsum)
```

```{r}
x <- data.frame(letter = LETTERS)
class(x[1,])
```

```{r}
x <- data.frame(letter = LETTERS)
class(x[1, 1, drop = FALSE])
```

```{r}
x <- data.frame(x = 1)
y <- data.frame(x = 2)
z <- rbind(x[1,1], y[1,1])
class(z)
```
```{r}
rep(TRUE, 5) & rep(FALSE, 5)
```

```{r}
rep(TRUE, 5) && rep(FALSE, 5)
```

```{r}
x <- c(1, 2, 3)
y <- c("A", "B", "C")
z <- c(x, y)
class(z)
```

```{r}
x <- c(1, 2, NA, 4, NA)
length(is.na(x))
```

```{r}
x <- c(1, NA, 3)
y <- c(NA, 2, NA)
all(is.na(x + y))
```

## Summary

By the end of this lesson, you should have feel touched up on your general R skills and you also should have seen some of the trickier parts of R.
Hopefully having seen the trickier parts of R will help later on down the road.

## Resources

Other good resources:

- One of our lessons on R: https://github.nceas.ucsb.edu/Training/R-intro
- Great, free book: http://r4ds.had.co.nz/
- DataCamp's R intro, https://www.datacamp.com/courses/free-introduction-to-r
Sources:
- Used https://www.rstudio.com/resources/cheatsheets/ to make sure I wasn't missing any basic stuff