---
title: "Ecological Metadata in R"
---

Time slot: 30 min

## Learning outcomes:

Upon completing this module, students will

- Have a basic understanding of scientific metadata and how it fits into research
- Have a basic familiarity with generating scientific metadata using the EML R package
- Have watched the instructor generate fairly complete metadata for an example dataset

## Introduction

Science metadata is:

- Machine-readable (usually XML)
- Standardized
- Usually really hard to generate

The R EML package aims to help us generate EML science metadata from within R.
The learning process here is two-fold:

- Learn the EML standard
- Learn how to use the EML R package

Both of these are relatively hard!

### The EML standard

EML covers lots of stuff, importantly:

- Proper citation of your dataset
- Who is involved with the dataset and how
- Coverage (temporal, spatial, taxonomic)
- Methodological information
- Documentation on files and their formats

## Generating an EML record from scratch

This is modeled after this vignette inside the EML package: https://github.com/ropensci/EML/blob/master/vignettes/creating-EML.Rmd

### The data we're going to document

A fairly simple dataset that we didn't 
```{r}
data("iris")
head(iris)
summary(iris)
```

### Getting the information in order

First we load the EML package:

```{r}
library(EML)
```

The easiest way to create an EML record from scratch is to get the information into R first, then create the EML record with that information.

So let's start with the title and abstract:

```{r}
title <- "Edgar Anderson's Iris Data"
abstract <- as(set_TextType("./abstract.md"), "abstract")
intellectualRights <- "This work is licensed under a Creative Commons Attribution 4.0 International License."
```
TODO: Set keywords

```{r}
keywordSet <-
  c(new("keywordSet",
        keyword = c("iris",
                    "ra fisher",
                    "setosa",
                    "virginica",
                    "versicolor")))
keywordSet
```
TODO: Set creator/contact, maybe another role

```{r}
edgar <- as.person("Edgar Anderson <edgaranderson@iris.net>") # Fake email
creator <- as(edgar, "creator")
contact <- as(edgar, "contact")
```

TODO: Set methods from file
```{r}
methods <- set_methods("methods.md")
methods
```


TODO: Set coverage

```{r}
coverage <- 
  set_coverage(begin = '1936-01-01', end = '1936-12-31', # Fake tempporal information
               sci_names = c("Iris setosa", "Iris versicolor", "Iris virginica"),
               geographicDescription = "Gaspé Peninsula", # Approximated spatial coverage
               west = -65.75, east = -65.75,
               north = 48.66, south = 48.66)
coverage
```

TODO: Set a dataTable

```{r}
attributes <- read.csv("attributes.csv")

species_codes <- c("setosa" = "Iris setosa",
                   "virginica" = "Iris virginica",
                   "versicolor" = "Iris versicolo")

factors <- data.frame(attributeName = "Species",
                      code = names(species_codes),
                      definition = species_codes)

attributeList <- set_attributes(attributes, 
                                factors,
                                col_classes = c("numeric",
                                                "numeric",
                                                "numeric",
                                                "numeric",
                                                "factor"))
attributeList
```

```{r}
write.csv(iris, row.names = FALSE, "iris.csv")
physical <- set_physical("iris.csv", 
                         size = as.character(file.size("iris.csv")),
                         authentication = digest::digest("iris.csv", algo = "md5", file = TRUE),
                         authMethod = "MD5")
physical
```

```{r}
dataTable <- new("dataTable",
                 entityName = "iris.csv",
                 entityDescription = "Edgar Anderosn's Iris data exported from R",
                 physical = physical,
                 attributeList = attributeList)
```


```{r}
dataset <- new("dataset",
               title = title,
               creator = edgar,
               pubDate = "2017",
               intellectualRights = intellectualRights,
               abstract = abstract,
               keywordSet = keywordSet,
               coverage = coverage,
               contact = contact,
               methods = methods,
               dataTable = dataTable)

eml <- new("eml",
           packageId = paste0("urn:uuid", uuid::UUIDgenerate()),
           system = "uuid",
           dataset = dataset)
```

```{r}
write_eml(eml, "eml.xml")
eml_validate("eml.xml")
```

TODO: Write it out
## Resources

https://knb.ecoinformatics.org/#external//emlparser/docs/index.html
https://knb.ecoinformatics.org/#external//emlparser/docs/eml-2.1.1/index.html