---
title: "Ecological Metadata in R"
---

Time slot: 30 min

## Learning outcomes:

Upon completing this module, students will

- Have a basic understanding of scientific metadata and how it fits into research
- Have a basic familiarity with generating scientific metadata using the EML R package
- Have watched the instructor generate fairly complete metadata for an example dataset

## Introduction

Science metadata is:

- Machine-readable
- Standardized
- Usually really hard to generate

The R EML package aims to help us generate EML science metadata from within R.
The learning process here is two-fold:

- Learn the EML standard
- Learn how to use the EML R package

Both of these are relatively hard!

### The EML standard

EML covers lots of stuff, importantly:

- Proper citation
- Who is involved with the dataset and how
- Coverage (temporal, spatial, taxonomic)
- Methods
- Docs on files and their formats

## Generating an EML record from scratch

Modeled after this vignette inside the EML package: https://github.com/ropensci/EML/blob/master/vignettes/creating-EML.Rmd

```{r}
library(EML)
```

First we start out with a new document:

```{r}
doc <- new("eml",
           dataset = new("dataset",
                         title = "My title"))
doc
```

TODO: Set title
TODO: Set abstract from file
TODO: Set keywords
TODO: Set creator/contact, maybe another role
TODO: Set methods from file
TODO: Set coverage
TODO: Set a dataTable
TODO: Write it out
## Resources

https://knb.ecoinformatics.org/#external//emlparser/docs/index.html
https://knb.ecoinformatics.org/#external//emlparser/docs/eml-2.1.1/index.html