---
title: "Ecological Metadata in R"
time_slot: 30 minutes
---

## Introduction

Science metadata is:

- Standardized
- Machine-readable (usually XML)
- Usually really hard to generate

The R EML package aims to help us generate EML science metadata from within R.
The learning process here is two-fold:

- Learn the EML standard
- Learn how to use the EML R package

Both of these are relatively hard!

## Learning outcomes:

Upon completing this module, students will

- Have a basic understanding of scientific metadata and how it fits into research
- Have a basic familiarity with generating scientific metadata using the EML R package
- Have watched the instructor generate fairly complete metadata for an simple example dataset

## Lesson

### The EML standard

EML covers lots of stuff, importantly:

- Proper citation of your dataset
- Who is involved with the dataset and how
- Coverage (temporal, spatial, taxonomic)
- Methodological information
- Documentation on files and their formats

## Generating an EML record from scratch

This is modeled after this vignette inside the EML package: https://github.com/ropensci/EML/blob/master/vignettes/creating-EML.Rmd

### The dataset

As an example, let's create an EML record for the `iris` dataset that comes with the `ggplot2` package.

```{r}
data("iris") # requires ggplot2
head(iris)
summary(iris)
```

### The metadata

First we load the EML package:

```{r}
library(EML)
```

The easiest way to create an EML record from scratch is to get the information into R first, then create the EML record with that information.

So let's start with the title and abstract:

```{r}
title <- "Edgar Anderson's Iris Data"
```

The easiest way to set an abstract is to create a separate Markdown file which lets us use rich formatting:

```{r}
abstract <- as(set_TextType("./abstract.md"), "abstract")
```

Though not required, including licensing information is a crucial step in metadata authoring.
Let's use the Create Commons Attribute license which is a permissive license.

```{r}
intellectualRights <- "This work is licensed under a Creative Commons Attribution 4.0 International License."
```

Every dataset should have a publication date.
I can guess at the most appropriate publication date from the ?iris help page.

```{r}
pubDate <- "1935"
```

#### Keywords

Search systems often take advantage of keywords to make it easier to find what you're looking for and find related datasets.

```{r}
keywordSet <-
  c(new("keywordSet",
        keyword = c("iris",
                    "ra fisher",
                    "setosa",
                    "virginica",
                    "versicolor")))
keywordSet
```

#### Parties

Every EML record needs to have a creator and a contact set.
The creator is the party or parties (e.g., person, organization) that should be cited when giving credit for the dataset.

```{r}
edgar <- as.person("Edgar Anderson <edgaranderson@iris.net>") # Fake email
creator <- as(edgar, "creator")
contact <- as(edgar, "contact")
```

#### Methods

We don't have detailed methods for this dataset so we'll make something up.
The easiest way to get methods into an EML record is to create a separate Markdown file which lets us get rich formatting.

```{r}
methods <- set_methods("methods.md")
methods
```

#### Coverage

We also don't have detailed coverage information but we can fill some things in from a bit of research.

```{r}
coverage <- 
  set_coverage(beginDate = '1936-01-01', 
               endDate = '1936-12-31', # Fake tempporal information
               sci_names = c("Iris setosa", "Iris versicolor", "Iris virginica"),
               geographicDescription = "Gaspé Peninsula", # Approximated spatial coverage
               westBoundingCoordinate = -65.75, 
               eastBoundingCoordinate = -65.75,
               northBoundingCoordinate = 48.66, 
               southBoundingCoordinate = 48.66)
coverage
```

#### Attributes

Attributes are one of the more powerful parts of EML.
We can describe, in very specific detail, the meaning of the tabular data we're documenting.
A lot of information is required to sufficiently describe datasets so we'll have to enter in a fair bit of information.
The easiest way to do that is to create a separate CSV file with a set of columns that the EML package is looking for and bring it in as a `data.frame`.

```{r}
attributes <- read.csv("attributes.csv")

# For the Species column, we need to define the values as codes and we need
# to tell EML what they mean
species_codes <- c("setosa" = "Iris setosa",
                   "virginica" = "Iris virginica",
                   "versicolor" = "Iris versicolo")

factors <- data.frame(attributeName = "Species",
                      code = names(species_codes),
                      definition = species_codes)

attributeList <- set_attributes(attributes, 
                                factors,
                                col_classes = c("numeric",
                                                "numeric",
                                                "numeric",
                                                "numeric",
                                                "factor"))
attributeList
```

#### Entities

We've described the attributes (columns) for `iris.csv` but we haven't describe `iris.csv` itself.
In EML, files like this are called `entities` and `entities` contain information about their file formats and more.

```{r}
write.csv(iris, row.names = FALSE, "iris.csv")
physical <- set_physical("iris.csv", 
                         size = as.character(file.size("iris.csv")),
                         authentication = digest::digest("iris.csv", algo = "md5", file = TRUE),
                         authMethod = "MD5")
physical
```

Because `iris.csv` is tabular, we create an `entity` of type `dataTable`:

```{r}
dataTable <- new("dataTable",
                 entityName = "iris.csv",
                 entityDescription = "Edgar Anderosn's Iris data exported from R",
                 physical = physical,
                 attributeList = attributeList)
```

Note that the `attributeList` we created before gets entered directly into the `dataTable` entity.

#### Create the `eml` object

Now that we have everything all entered into R, we can create the 
```{r}
dataset <- new("dataset",
               title = title,
               creator = edgar,
               pubDate = pubDate,
               intellectualRights = intellectualRights,
               abstract = abstract,
               keywordSet = keywordSet,
               coverage = coverage,
               contact = contact,
               methods = methods,
               dataTable = dataTable)

eml <- new("eml",
           packageId = paste0("urn:uuid", uuid::UUIDgenerate()),
           system = "uuid",
           dataset = dataset)
```

#### Save and validate

Now that our `eml` object is created, we can save it:

```{r}
write_eml(eml, "eml.xml")
```

We should also validate the file:

```{r}
eml_validate("eml.xml")
```

## Summary

## Resources

- https://knb.ecoinformatics.org/#external//emlparser/docs/index.html
- https://knb.ecoinformatics.org/#external//emlparser/docs/eml-2.1.1/index.html
- https://github.com/ropensci/EML