---
title: "Regular Expressions"
author: "Bryce Mecum"
---

Time slot: 30 minutes

Learning outcomes:
  
Students should:
  
- Understand when regular expressions are appropriate
- Have an introductory-level awareness of regular expression syntax
- Have some experience executing and working with regular expressions in R

## Overview and Examples

Regular expressions are a fantastic tool for filtering and even extracting information out of strings of characters such as site codes, titles, or even entire documents. 
Regular expressions follow a custom syntax that we'll need to learn but they are worth learning because:

- Regular expressions can do things other methods cannot
- Regular expressions can be used with many other languages and tools so it's a learn-once, use-everywhere kind of tool

Let's start off with a simple example of where simpler methods won't work and see how regular expressions can be used to get what we need done:

Let's say we just received some data we need to analyze and we find this:

```{r}
site_data <- read.csv("site_data.csv", stringsAsFactors = FALSE)
site_data
```

It looks like the author of the dataset mixed the year of measurements, site code (e.g., A, CCCC, etc.), and some sub-site code (e.g., 1, 2, 3, etc.) into a single column.
If we wanted to, for example, calculate mean temperature by site, we'd need to split these up somehow into separate columns.
How could we go about this?
We could start with `substr` which lets us slice a string by its indices:

```{r}
substr(site_data$x, 5, 7)
```

But we'd quickly find that, because the number of characters in the site code varies from one to four, we can't extract just the site code. 
These are the types of problems where regular expressions come in handy.

Before we start, we're going to use the `str_view_all` function from the `stringr` package which shows a nice display of the result of executing a regular expression against our strings.
In real use, we would use another function to actually get and work with the result.

```{r}
library(stringr)
str_view_all(site_data$x, "[A-Z]+")
```

The expression we used above, `[A-Z]+`, is equivalent to asking for the first consecutive run of the letters A-Z in the entire string of characters.

### Another example

Data often come to us in strange forms and, before we can even begin analyzing the data, we have to do a lot of work to sanitize what we've been given.
An example, which I just got the other week were temporal data with dates formatted like this:

```{r}
dates <- c("1July17",
           "02July2017",
           "3July17",
           "4July17")
```

and so on like that.
Do you see how the day of the month and year are represented in different ways through the series?
If we want to convert these strings into `Date` objects for further analysis, we'll to do some pre-cleaning before we can do that conversion.
Regular expressions work great here.

```{r}
str_match_all(dates, "(\\d{1,2})([A-Za-z]+)(\\d{2,4})")
```

## Overview of Regular Expressions

Regular expressions can match things literally, e.g.,

```{r}
str_detect("apple", "apple")
```

but they also support a large set of special characters:

- `.`: Match any character

```{r}
words <- c("apple", "banana", "fork")
str_view_all(words, "a.")
```

If you actually want to match a period and not any character, you have to do what's called escaping:

```{r}
str_view_all("apples are a fruit and my favorite fruit is an apple.", "apple\\.")
```

See how that regular expression only matched the apple with the period at the end?

- `[]`: Match any character in this set

```{r}
words <- c("apple", "banana", "fork")
str_view_all(words, "[aeiou]")
```


- `[^]`: Match any character *not* in this set

```{r}
words <- c("apple", "banana", "fork")
str_view_all(words, "[^aeiou]")
```

- `\s` & `\S`: Match any whitespace (e.g., ` `, `\t`)

```{r}
words <- c("apple", "banana", "fork")
str_view_all(words, "\\s") # Note the double \\ before the s. This is an R-specific thing.
                           # many of our special characters must be preceded by a \\
str_view_all(words, "\\S")

pies <- c("lemon meringue pie", "apple pie", "strawberry rhubarb pie")
str_view_all(pies, "\\s")
```


- `\d` & `\D`: Match any digit, equivalent to `[0-9]`

```{r}
x <- c("apple", "1234", "banana")
str_view_all(x, "\\d")
```

- `\w` & `\W`: Match any word character, equivalent to `[A-Za-z0-9_]`

```{r}
x <- c("apple", "1234", "banana")
str_view_all(x, "\\w")
```

We can also specify how many of a particular character or class of character to match:

- `?` Optionality / 0 or 1

Say we want to get just the phone numbers out of this vector but we notice that the phone numbers take on some different formats:

```{r}
phone_numbers <- c("219 733 8965", "apple", "329-293-8753 ", "123banana", "595.794.7569", "3872876718")
str_view_all(phone_numbers, "\\d\\d\\d[ \\.-]?\\d\\d\\d[ \\.-]?\\d\\d\\d\\d")
```

The above regular expression matches the number parts of the phone numbers, which can be separated by zero or one space (` `), `.`, or `-`.

- `+` 1 -> infinity

We can use the `+` expression to find words with one or more vowels:

```{r}
words <- c("apple", "crypt", "banana", "pygmy", "boo")
str_view_all(words, "[aeiuo]+")
```

- `*` 0 -> infinity

and the `*` is _zero or more_`.

```{r}
numbers <- c("0.2", "123.1", "547")
str_view_all(numbers, "\\d*\\.?\\d*")
```


- `()`: Grouping

One of the most powerful parts of regular expressions is grouping.
Grouping allows us to split up our matched expressions and do more work with them.
For example, we can create match the city and state in a set of addresses, splitting it into components:

```{r}
addresses <- c("Santa Barbara, CA", "Seattle, WA", "New York, NY")
str_view_all(addresses, "([\\w\\s]+), (\\w+)")
```

Once we use groups, `()`, we can also use back references to work with the result:

```{r}
str_replace(addresses, "(\\w+), (\\w+)", "City: \\1, State: \\2")
```

- `^` & `$`

It can also be really useful to make a say something like "strings that start with a capital letter" or "strings that end with a period":

```{r}
possible_sentences <- c(
  "This might be a sentence.",
  "So. Might. this",
  "but this could maybe not be?",
  "Am I a sentence?",
  "maybe not",
  "Regular expressions are useful!"
)
# ^ specifies the start so ^[A-z] means "starts with a capital letter""
str_detect(possible_sentences, "^[A-Z]")
# We can also do "ends with a period"
str_detect(possible_sentences, "\\.$")
# We can put them together:
str_detect(possible_sentences, "^[A-Z].*[\\.\\?!]$")
```


## Finish out our example together

Now that we've gone over some basics of regular expressions, let's finish our example by splitting the various components of column `x` into a `year`, `site`, and `sub_site` column:

```{r}
```

# Common R functions that use regular expressions

- Base R
  - `grep`
  - `gsub`
  - `strsplit`
- `stringr` package
  - `string_detect`
  - `string_match`
  - `string_replace`
  - `string_split`

## More

- Have the group figure out you can put * onto [].
- Have the group str_split on fixed chars and a regex

## Resources

- [https://regex101.com/](https://regex101.com/) (This is the site I usually use)
- [https://docs.microsoft.com/en-us/dotnet/standard/base-types/regular-expression-language-quick-reference](https://docs.microsoft.com/en-us/dotnet/standard/base-types/regular-expression-language-quick-reference)
- [https://developer.mozilla.org/en-US/docs/Web/JavaScript/Guide/Regular_Expressions](https://developer.mozilla.org/en-US/docs/Web/JavaScript/Guide/Regular_Expressions)

## Appendicies

Here's the code I used to generate the fake site_data `data.frame` above.

```{r, eval=FALSE, echo=FALSE, include=FALSE}
# Fake data generation code
site_data <- data.frame(year = rep(seq(2000, 2009), 4))
site_data$site <- sample(c("A", "AA", "B", "CCCC"), nrow(site_data), replace = TRUE)
site_data$subsite <- sample(c(1,2,3,4), nrow(site_data), replace=TRUE)
site_data$temp_c <- runif(nrow(site_data), 0, 100)
site_data$site_code <- paste(site_data$year, site_data$site, site_data$subsite, sep="-")
site_data <- site_data[,c("site_code", "temp_c")]
names(site_data) <- c("x", "temp_c")
write.csv(site_data, file = "site_data.csv", row.names = FALSE)
```

