---
title: "Uploading Data to DataONE"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Uploading Data to DataONE}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

## Introduction
  
This document describes how to use the *dataone* R package to upload data to DataONE.

The *dataone* R package provides methods to enable R scripts to interact with DataONE Coordinating Nodes (CN) and Member Nodes (MN), to search for, download, upload and update data and metadata. The *dataone* R package takes care of 
the details of calling the  corresponding DataONE web service on a DataONE node. For example, the *dataone* 
`createObject` R method calls the DataONE web service [MNStorage.create()](https://purl.dataone.org/architecture/apis/MN_APIs.html#MNStorage.create) that uploads a dataset to a DataONE MN. 

## Logging in
Before uploading any data to a DataONE MN, it is necessary to obtain a DataONE user identity that will
be provided with each request to upload or update data. The method that DataONE uses
requires an *authentication token*, which is a character string, to be provided during upload. This token can be retrieved by logging into the DataONE network and copying the token into your R session.  

### Obtain an ORCID
ORCID is a researcher identifier that provides a common way to link your researcher identity to your articles and data.  An ORCID is to a researcher as a DOI is to a research article.  To obtain an ORCID, register at https://orcid.org.

### Log in to DataONE
We will be using a test server, so login and retrieve your token at https://search-stage-2.test.dataone.org.

Once you are logged in, navigate to your Profile Settings, and locate the "Authentication Token" section, and then copy the token for R to your clipboard.

## Modify metadata
Next, modify the metadata file associated with the package to set yourself as the owner.  This will help us differentiate the test data later.  Open the `strix-pacific-northwest.xml` file in RStudio, and change the `givenName` and `surName` fields at the top to your name.

```{r}
library(EML)
library(arcticdatautils)

# Load the EML file into R
doc <- read_eml("strix-pacific-northwest.xml")

# Change creator to us
doc@dataset@creator <- c(eml_creator("Bryce", "Mecum", email = "mecum@nceas.ucsb.edu"))

# Change abstract to the better one we wrote
doc@dataset@abstract <- as(set_TextType("better-abstract.md"), "abstract")

# Save it back to the filesystem
write_eml(doc, "strix-pacific-northwest.xml")
```

## Uploading A Package Using `uploadDataPackage`
Datasets and metadata can be uploaded individually or as a collection. Such a collection, whether contained in local 
R objects or existing on a DataONE repository, will be informally referred to as a `package`. The steps necessary to
to prepare and upload a package to DataONE using the `uploadDataPackage` method 
will be shown. A complete script that uses these steps is shown here.

In the first section, we create a 'DataPackage as a container for our data and metadata and scripts:

```{r}
library(dataone)
library(datapack)
library(uuid)

dp <- new("DataPackage")
dp
```

We then add a metadata file, data file, R script and output data file to this package:
```{r}
emlFile <- "strix-pacific-northwest.xml"
metadataObj <- new("DataObject", format="eml://ecoinformatics.org/eml-2.1.1", filename=paste(getwd(), emlFile, sep="/"),
                   suggestedFilename=basename(emlFile))
dp <- addData(dp, metadataObj)

sourceData <- "sample.csv"
sourceObj <- new("DataObject", format="text/csv", filename=paste(getwd(), sourceData, sep="/"), suggestedFilename=basename(sourceData))
dp <- addData(dp, sourceObj, metadataObj)

progFile <- "filterSpecies.R"
progObj <- new("DataObject", format="application/R", filename=paste(getwd(), progFile, sep="/"), mediaType="text/x-rsrc",
                suggestedFilename=basename(progFile))
dp <- addData(dp, progObj, metadataObj)

outputData <- "filteredSpecies.csv"
outputObj <- new("DataObject", format="text/csv", filename=paste(getwd(), outputData, sep="/"), suggestedFilename=basename(outputData))
dp <- addData(dp, outputObj, metadataObj)

myAccessRules <- data.frame(subject="http://orcid.org/0000-0003-0077-4738", permission="changePermission") 

dp
```

Finally, we upload the package to the Testing server for the KNB.
```{r, eval=FALSE}
d1c <- D1Client("STAGING2", "urn:node:mnTestKNB")
packageId <- uploadDataPackage(d1c, dp, public=TRUE, accessRules=myAccessRules, quiet=FALSE)
```

This particular package contains the R script `filterSpecies.R`, the input file `sample.csv` that was read by
the script and the output file `filteredSpecies.csv` that was created by the R script, which was run at a
previous time.  

You can now search for and view the package at https://dev.nceas.ucsb.edu