---
title: "Testing R Code and Analysis"
time_slot: 30 minutes
---

When we write R code to do our analyses, we're usually making use of other peoples' functions.
But sometimes we have a need to write our own functions and, when we do this, it is a great idea to test them to make sure they work as we expect.

As an example, we'll create a new R package that provides a single function to calculate Z scores for a vector of numbers.
A Z score tells us how many standard deviations a value is from the mean of a set of values.
Mathematically, it is defined as $z = (X - μ) / σ$ where $X$ is the value, $µ$ is the mean and $σ$ is the standard deviation.
In reality, this is the `scale` function in R but we're going to write it for ourselves as a simple example.

- Step 1: Get set up
  - Use already-created package if we get there in previous lesson
  
```{r}
install.packages(c("testthat", "devtools"))
```

```{r}
devtools::use_testthat()
```

rename hello.R to z_score.R

replace hellow.R with:

```{r}
z_score <- function(x) {
  x
}
```

source it
runit in console

```{r}
z_score <- function(x) {
  (x - mean(x)) / sd(x)
}
```

- Create tests/testthat/test_z_score.R

```
context("z_score")
```
      
      - Each test file should start with test and end in .R or .r
- Step 2: Test our code and analyis
  - Basic code testing analysis
  - Show a model function being tested